Likelihoods for a general class of ARGs under the SMC

### 2025-04-29 review-responses.tex

\begin{point}{\revref} % 300:
 for this application, despite retaininig fewer nodes and edges.
\end{point}

\reply{
We would prefer to keep this sentence as it is, as ``retaining" here could 
be confusing --- the calculation doesn't keep or retain nodes and edges, that is a 
property of the input ARG.
}

\begin{point}{\revref{} and subsequent paragraph} % 306-316:
 Is this number of polytomies too small and is then the impact on the
likelihood small? It would be good to do one more extreme example for
supplement to ensure this. This is the point where you can strengthen your
paper the most.
\end{point}

\reply{
We have added a supplemental figure which explores the effects of deleting
up to 75\% of the internal nodes, and the overall picture remains the 
same. Remarkably, this seems to improve
performance in the fully-simplified case. Hopefully this will spark 
some interest in follow-up studies.
}

@@

\begin{point}{Fig \ref{fig:lik-surface} caption:}
 lines overlap hence we cannot see the full red line.
\end{point}

\reply{
We have added small red dots on the red line to make it legible.
}

### 2025-04-29 paper.tex
Add new supplemental figure.

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has only a very minor effect on location of the minimum.

@@

>>>>

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has only a very minor effect on location of the minimum.
Supplemental Fig.~S1 explores more extreme examples in which we remove
25\%, 50\% and 75\% of the internal nodes in these ARGs, and demonstrates 
a remarkable robustness to the loss of the majority of the internal
nodes in these examples. Interestingly, the loss of 75\% of the 
internal nodes  in the fully simplified ARG 
(1841 of 2455, leading  to an average of 27 polytomies per tree)
appears to improve performance, substantially 
reducing the bias away from the true parameter value.

@@

\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[!ht]
\centering
\includegraphics{figures/likelihood_surface_suppl}
    \caption{Log likelihood-ratio curves for three different ARGs 
    with increasing fractions of polytomies.
    The base ARGs are as described in Fig.~3, but here we remove 
    25\%, 50\% and 75\% of the internal nodes to create polytomies.
    The true value of recombination parameter is indicated by the 
    dashed line, as in Fig.~3.
    \label{fig:likelihood_suppl}}
\end{figure}
